Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment

### 2021-09-17 manuscript.Rmd
Re-run COVID severity model, correct to first only.

-----
\newpage
<P style = "page-break-before: always">

@@

* __Background__: SARS-CoV-2 RNA has been described to contaminate surfaces within healthcare settings. However, little is known regarding the role of distance and time in surface contamination in a clinical setting.

@@

* __Results__: Probability of detecting SARS-CoV-2 RNA did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: odds ratio [OR] 67.77 (95% confidence interval [CI] 36.34 to 131.47, high-touch elevated: OR 7.39 (95% CI 4.31 to 13.09)). Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory support had greater odds of  surface contamination (OR 1.76 (95% CrI 1.2 to 2.57)). While the probability of surface contamination decayed with prolonged hospitalization, probability of floor detection increased with time over the pandemic.

@@

 <!-- -----
\newpage
<P style = "page-break-before: always">

@@

We have previously applied spatial mapping and systematic sampling to define how multidrug-resistant organisms cluster in the hospital environment [site pending ICHE manuscript]. We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.

-----  
\newpage
<P style = "page-break-before: always">

@@

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA. (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface. Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).

* __Causal Model__: We hypothesized that SARS-CoV-2 contamination would decay with physical distance from patient and with time from COVID-19 diagnosis. We also hypothesized that high-touch objects would be more frequently contaminated with SARS-CoV-2 RNA than low-touch objects.

@@

* __Characteristics of subjects hospitalized for COVID-19 during the second (fall 2020) wave__: We collected `r nrow(dat)` specimens from `r nrow(count(dat,subject_room_id))` unique subject-room pairs, comprising `r nrow(count(dat,redcap_id))` unique subjects (i.e., `r nrow(filter(count(distinct(select(dat,subject_room_id,redcap_id)),redcap_id),n>1))` subjects were transferred at least once during the observation period). The median number of sampling events from each subject-room pair was `r round(mean(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)` (SD `r round(sd(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)`). **Table 1** summarizes the environmental surfaces sampled, whether they were high-touch or low-touch, the proportion of each specimen type from which SARS-CoV-2 RNA was detected by RT-qPCR, and the SARS-CoV-2 copy numbers measured. We did not observe significant differences in copy numbers across the surface types, but the proportion with detectable SARS-CoV-2 did vary widely.

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of MDRO organisms [@Kelly2021], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% round(.,2)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

* __SARS-CoV-2 RNA Environmental Contamination Decays During Prolonged Hospitalization__: We evaluated the effect of time from COVID-19 diagnosis on the detection of SARS-CoV-2 RNA in the hospital environment via a multivariable logistic regression model relating time from COVID-19 diagnosis, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection. We found that the probability of detection decreased as time from COVID-19 diagnosis increased, with OR `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q97.5_exp) %>% round(.,2)`) per day since COVID-19 diagnosis. The most prominent effects were observed on high-touch elevated surfaces and floor surfaces (**Figure 1B**).

* __SARS-CoV-2 Contamination of the Floor Increases with Pandemic Wave Duration__: To understand how subjects differed in the effect of time from COVID-19 diagnosis on SARS-CoV-2 contamination of the hospital environment, we re-fit the above model with the addition of random intercepts and slopes to account for subject-level differences. **Figure 2A** depicts the best estimate of the effect of time from COVID-19 diagnosis on the contamination in each subject's hospital room. In the figure, the lines representing each subject are colored by when they were enrolled during the course of the local, second-wave COVID-19 case surge. We observed that the contamination of high-touch elevated surfaces did not appear to relate to time from the start of the local case surge. However, contamination of floor surfaces appeared to be greater among subjects enrolled later in the course of the local case surge. To investigate this further, we added an interaction term to the model and found that each day since the start of the start of the local second wave increased the contamination of floor surfaces with OR `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Estimate_exp) %>% round(.,2)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q97.5_exp) %>% round(.,2)`) (**Figure 2B**).

* __SARS-CoV-2 Contamination of Increases with COVID-19 Disease Severity__: To understand how COVID-19 disease severity could impact environmental contamination with SARS-CoV-2 RNA, we categorized subjects based on their required oxygen support: `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[1]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[1]`) subjects who required no oxygen support, `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[2]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[2]`) subjects who required only nasal or facemask oxygen support, and `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[3]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(round(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[3]`) subjects who required significant (i.e., high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation) oxygen support. Adjusting for the site, time from COVID-19 diagnosis, and time from start of the local COVID-19 wave, we found that the use of significant oxygen support was associated with increased environmental SARS-CoV-2 RNA contamination, with OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% round(.,2)` ((95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% round(.,2)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% round(.,2)`)) relative to the no-oxygen-support reference group. **Figure 3** shows the predicted probabilities of SARS-CoV-2 environmental contamination as it relates to COVID-19 severity across all measured surfaces, adjusted for time from COVID-19 diagnosis and time from the start of the local COVID-19 wave.

-----  
\newpage
<P style = "page-break-before: always">

@@

We sought to define the impact of time and distance on SARS-CoV-2 surface contamination through longitudinal sampling of  within COVID-19 patient rooms. We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA. We had hypothesized that the probability of RNA detection would decrease with distance. This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021]. However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied. This is most likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and healthcare personnel (HCP) foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room. This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

@@

With increased clinical severity, specifically those who required high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation, we detected an increased odds of detecting surface RNA within rooms. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aeresol production associated with respiratory procedures. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021] in a large observational study, potentially supporting the contribution of respiratory care in surface contamination. Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients[@CDC2021].

@@

__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ .

-----  
\newpage
<P style = "page-break-before: always">

>>>>

-----

\newpage

@@

* __Background__: SARS-CoV-2 is known to contaminate surfaces within healthcare settings, but the spatial and temporal extent of environmental contamination has not been precisely defined. We sought to define SARS-CoV-2 contamination of different surface types and how contamination changes over time.

@@

* __Results__: The probability of detecting SARS-CoV-2 RNA did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors: `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated: OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)). Patients requiring high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation respiratory were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)). The probability of surface contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.

@@

 <!-- -----  
\newpage

@@

We have previously applied a systematic sampling and spatial modeling strategy to define how multidrug-resistant organisms cluster in the hospital environment [@Kelly2021-vm]. We now extend the methods we have developed to understand how SARS-CoV-2 contamination is distributed spatially, and how it changes over time.

-----

\newpage

@@

* __Study Design, Setting, and Population__: We performed a prospective cohort study at the Hospital of the University of Pennsylvania (HUP), an academic medical center in Philadelphia, PA, USA (IRB protocol #843273). Patients with COVID-19 were housed on multiple hospital units over the course of the study according to bed availability. All rooms underwent daily cleaning, according to hospital protocol. Environmental services (EVS) staff used disposable disinfectant wipes (quaternary ammonium disinfectant) for use on high-touch objects, limiting use to a single surface. Floors were cleaned using a microfiber mop and a chlorine-based disinfectant (sodium dichloroisocyanurate).

* __Causal Model__: We hypothesized that SARS-CoV-2 contamination would decay with physical distance from patient and with time from COVID-19 diagnosis. We also hypothesized that high-touch objects would be more frequently contaminated with SARS-CoV-2 RNA than low-touch objects, and that increased COVID-19 disease severity would be associated with greater environmental contamination.

@@

* __Characteristics of subjects hospitalized for COVID-19 during the second (fall 2020) wave__: We collected `r nrow(dat)` specimens from `r nrow(count(dat,subject_room_id))` unique subject-room pairs, comprising `r nrow(count(dat,redcap_id))` unique subjects (i.e., `r nrow(filter(count(distinct(select(dat,subject_room_id,redcap_id)),redcap_id),n>1))` subjects were transferred at least once during the observation period). The median number of sampling events from each subject-room pair was `r signif(mean(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)` (SD `r signif(sd(count(distinct(select(dat, subject_room_id, subject_room_day)), subject_room_id)$n, na.rm = TRUE),2)`). **Table 1** summarizes the environmental surfaces sampled, whether they were high-touch or low-touch, the proportion of each specimen type from which SARS-CoV-2 RNA was detected by RT-qPCR, and the SARS-CoV-2 copy numbers measured. We did not observe significant differences in copy numbers across the surface types, but the proportion with detectable SARS-CoV-2 did vary widely.

* __SARS-CoV-2 RNA Environmental Contamination Varies with Surface Type But Not Distance from Patient__: We evaluated the relationship between the distance from the patient's head of the bed and the probability of detecting SARS-CoV-2 on environmental surfaces. There was no significant relationship between distance and the probability of SARS-CoV-2 detection, in contrast to prior studies of multidrug-resistant bacterial organisms [@Kelly2021-vm], but the suface type was significantly associated with the probability of detection. Floor surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to elevated surfaces. High-touch elevated surfaces had an odds ratio of `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`) of SARS-CoV-2 detection, relative to low-touch elevated surfaces. **Figure 1A** depicts the results of a multivariable model relating distance from patient, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection.

* __SARS-CoV-2 RNA Environmental Contamination Decays During Prolonged Hospitalization__: We evaluated the effect of time from COVID-19 diagnosis on the detection of SARS-CoV-2 RNA in the hospital environment via a multivariable logistic regression model relating time from COVID-19 diagnosis, surface location, and whether the surface was high-touch to the probability of SARS-CoV-2 detection. We found that the probability of detection decreased as time from COVID-19 diagnosis increased, with OR `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Time Fixed" & param == "b_subject_covid_day") %>% pull(Q97.5_exp) %>% signif(.,3)`) per day since COVID-19 diagnosis. The most prominent effects were observed on high-touch elevated surfaces and floor surfaces (**Figure 1B**).

* __SARS-CoV-2 Contamination of the Floor Increases with Pandemic Wave Duration__: To understand how subjects differed in the effect of time from COVID-19 diagnosis on SARS-CoV-2 contamination of the hospital environment, we re-fit the above model with the addition of random intercepts and slopes to account for subject-level differences. **Figure 2A** depicts the best estimate of the effect of time from COVID-19 diagnosis on the contamination in each subject's hospital room. In the figure, the lines representing each subject are colored by when they were enrolled during the course of the local, second-wave COVID-19 case surge. We observed that the contamination of high-touch elevated surfaces did not appear to relate to time from the start of the local case surge. However, contamination of floor surfaces appeared to be greater among subjects enrolled later in the course of the local case surge. To investigate this further, we added an interaction term to the model and found that each day since the start of the start of the local second wave increased the contamination of floor surfaces with OR `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Time Pandemic Wave Adjusted" & param == "b_site_categoryFloor:total_study_day") %>% pull(Q97.5_exp) %>% signif(.,3)`) (**Figure 2B**).

* __SARS-CoV-2 Contamination of Increases with COVID-19 Disease Severity__: To understand how COVID-19 disease severity could impact environmental contamination with SARS-CoV-2 RNA, we categorized subjects based on their required oxygen support: `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[1]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(signif(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[1]`) subjects who required no oxygen support, `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[2]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(signif(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[2]`) subjects who required only nasal or facemask oxygen support, and `r dat_subject %>% count(covid_severity) %>% pull(n) %>% .[3]` (`r dat_subject %>% count(covid_severity) %>% mutate(proportion = paste0(signif(n / sum(n, na.rm = TRUE) * 100, 3),"%")) %>% pull(proportion) %>% .[3]`) subjects who required significant (i.e., high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation) oxygen support. Adjusting for the site, time from COVID-19 diagnosis, and time from start of the local COVID-19 wave, we found that the use of significant oxygen support was associated with increased environmental SARS-CoV-2 RNA contamination, with OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`) relative to the no-oxygen-support reference group. **Figure 3** shows the predicted probabilities of SARS-CoV-2 environmental contamination as it relates to COVID-19 severity across all measured surfaces, adjusted for time from COVID-19 diagnosis and time from the start of the local COVID-19 wave.

-----

\newpage

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient's bed did not inform the probability of detection of SARS-CoV-2 RNA. We had hypothesized that the probability of RNA detection would decrease with distance. This hypothesis was informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room [@Kelly2021-vm]. However, we found that SARS-CoV-2 detection did not vary significantly among any surface group as distance varied. This is most likely related to two factors 1) the relatively small size of patient rooms and 2) the physical spread of viral RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and healthcare personnel (HCP) foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient's room. This finding matches previous description of high probability of floor contamination relative to other surfaces [@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

@@

With increased clinical severity, specifically those who required high-flow, continuous or bilevel positive airway pressure, or mechanical ventilation, we detected an increased odds of detecting surface RNA within rooms. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021] in a large observational study, potentially supporting the contribution of respiratory care in surface contamination. Thus, while, we can only hypothesize the mechanism for greater odds of detection in these patients, this finding supports CDC guidance for prolonged isolation in critically ill patients [@CDC2021-gc].

@@

__Table 1: Hospital environment features sampled and proportion with detectable SARS-CoV-2 RNA.__ For each environmental surface type, whether the surface was considered high-touch or low-touch, the proportion of surfaces positive for SARS-CoV-2 RNA by RT-qPCR, and the median (IQR) SARS-CoV-2 copy numbers measured are reproted.

-----

\newpage
